prolfqua - An R-package for Proteomics Differential Expression Analysis.

### 2022-05-05 inst/MyArticle/paper.Rmd
some additions to the supplment

If there are no observation in a group, we assume that there are no observations because of the limit of detection (LOD). Furthermore, we assume that the variance of the protein is constant in all the conditions.
We impute the mean estimate of the group by using the average of $1\%$ smallest group mean intensities in the experiment. To estimate the variance in the condition, we use the pooled variance based on all conditions:

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%'}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()

@@

@@

To ease the usage barriers of the R package to users not being so familiar with statistics and R programming, we integrated some workflows into our Local Laboratory Information Management System (LIMS) [b-fabric](https://fgcz-bfabric.uzh.ch/) @bfabric.
This integration enables our users to select the input and basic settings in a graphical user interface. As an output, the user receives a report including the R markdown file and all necessary input files to reprocess the analysis on their devices.
In this way, we also support the reproducible science efforts of the community.

@@

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the pAUC, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice. First, estimate protein abundances from peptide abundances using a robust or nonparametric regression method. Second, use linear models as long as they suffice to model your experimental design because they show good performance in all categories. Third, use mixed effect models only if you have to, e.g., repeated measurements and if the sample sizes are large. If you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change.

The r-package proDA and prolfqua model the missing data directly, performing similarly, while MSstats imputes the data using an accelerated failure model. MSstats estimates group differences for fewer proteins and does not achieve a higher pAUC score despite imputation. Furthermore, \@ref(fig:FDRfdp) Panel C shows that when using MSstats the false proportion of false discoveries might be very high even when filtering for a low FDR.

In summary, `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data, report results, and benchmark MS software and statistical methods. We provide more details at the website https://github.com/fgcz/prolfqua/.

@@

## Creating a prolfqa configuration

The following code demonstrates how an analysis of MSFragger ouput can be performed using prolfqua.

@@

## Comparing benchmark results

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation, but using only the protein where inference could be performed for both models. We see that the fold change estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy.

@@

```{r proDAAllBenchmarks}
proda <- allBenchmarks$benchmark_proDA$data()
imp <- allBenchmarks$benchmark_imputation$data()

>>>>

If there is no observation in a group, we assume that the protein was unobserved because the protein abundance is below the limit of detection (LOD) of the MS instrument. We estimate the detection limit as follows: if a protein is quantified just in a single sample of a treatment group, then it is at the instrument's detection limit. We compute the mean of those abundances and use it as the group mean if a protein is absent in a treatment group.

When computing differences among two groups $a$ and $b$, we will use the group mean $\bar{a}$ estimated from the data or the detection limit $a_d$ if no observations are present in the group. If for instance there are no observations in group $b$, then, $\Delta_{a,b} = \bar{a} - b_{dl}$ if $a > b_{dl}$ or $\Delta = 0$ if $a < b_{dl}$.

Furthermore, we assume that the variance of the protein is constant in all the conditions. To estimate the variance in the condition, we use the pooled variance based on all conditions:

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 6}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()

@@

The r-package proDA and prolfqua model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, MSstats do not estimate group differences for more proteins and do not achieve a higher pAUC score than prolfqua. Furthermore, \@ref(fig:FDRfdp) Panel C shows that when using MSstats the proportion of false discoveries might be very high even when filtering for a low FDR.

@@

To ease the usage barriers of the R package to users not being so familiar with statistics and R programming, we integrated some workflows into our Local Laboratory Information Management System (LIMS) [b-fabric](https://fgcz-bfabric.uzh.ch/) @bfabric.
This integration enables our users to select the input and basic settings in a graphical user interface. As an output, the user receives a report including the R markdown file and all necessary input files to reprocess the analysis using R on their devices. In this way, we also support the reproducible science efforts of the community.

@@

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the pAUC, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice. First, estimate protein abundances from peptide abundances using a robust or nonparametric regression method. Second, use linear models as long as they suffice to model your experimental design because they show good performance in all categories. Third, use mixed effect models only if you have to, e.g., repeated measurements and if the sample sizes are large. If you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. Last but not least, do not impute missing data but model it instead.

In summary, `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data, report results, and benchmark MS software and statistical methods. We provide more details at the website https://github.com/fgcz/prolfqua/. 
This document was created using Rmarkdown, and all the code need to replicate the document and the presented benchmarks is available at : https://github.com/wolski/prolfquabenchmark/.

@@

## Creating a prolfqua configuration

The following code demonstrates how an analysis of MSFragger output can be performed using prolfqua.
First, a tidy table with protein abundances is generated by reading protein abundances reported by MSFragger in the combined_protein.tsv file.
The samples are annotated using the data stored in the annotation.xlsx file. Next, an _AnalysisTableAnnotation_ R6 object is created. 
Since bottom-up proteomics data is hierarchical a protein has peptides, which might be modified and create precursors. To describe this hierarchy, AnalysisTableAnnotation has a hierarchy field that stores a list with an entry for each hierarchy level. Since combined_portein.tsv only holds protein level data, the hierarchy list has only one element, and we need to specify which column does contain the protein identifiers.
We also need to specify which column contains the protein abundances we want to use for the data analysis.
Finally, we have to specify which columns contain the explanatory variables of the analysis. The AnalysisTableAnnotation has the field factors, a list with as many entries as explanatory variables. Here we have only one explanatory variable, the dilution, specified in the column 'sample.'

@@

## Comparing benchmark results for proDA and prolfqua

Here we compare the benchmark results for the `r BiocStyle::Biocpkg("proDA")` model and the linear model with variance moderation, but using only the protein where inference could be performed for both models. We see that the fold change estimates obtained with the `r BiocStyle::Biocpkg("proDA")` package show better diagnostic accuracy. Figure \@ref(fig:proDAAllBenchmarks), compares the fold change estimates obtained using both methods. However, the $t-statistics$ or $p-values$ show slightly worse diagnostic accuracy.

@@

```{r proDAAllBenchmarks, fig.cap="Comparison of fold change estimates produced by proDA and using the missing data modelling method implemented in prolfqua."}
proda <- allBenchmarks$benchmark_proDA$data()
imp <- allBenchmarks$benchmark_imputation$data()
imp <- allBenchmarks$benchmark_ProtModerated$data()
